\section{GMT Visualisation Tools}
There are many tools for visualising netCDF files. In particular \href{http://meteora.ucsd.edu/~pierce/ncview\_home\_page.html}{\texttt{ncview}} and \href{http://www.epic.noaa.gov/java/ncBrowse/}{\texttt{ncBrowse}} are usefule viewers. We also provide a number of Python programs and classes using \href{http://gmt.soest.hawaii.edu/}{GMT} which produce postscript images of the GLIMMER netCDF files.

The visualisation tools are based on GMT programs written in Python. In order to be able to use these it is necessary to have a few extra packages:
\begin{itemize}
\item \href{http://www.pfdubois.com/numpy/}{{\bf Numerical Python}}: Numerical Python extends python to be able to work on gridded data.
\item \href{http://starship.python.net/~hinsen/ScientificPython/}{{\bf Scientific Python}}: Scientific Python provides access to netCDF files. Make sure you enable the netCDF specific modules.
\end{itemize}

The visualisation tools are based on \texttt{PyGMT} which you can get from NeSCForge
{\small
\begin{verbatim}
cvs -d:pserver:anonymous@forge.nesc.ac.uk:/cvsroot/pygmt login
cvs -z3 -d:pserver:anonymous@forge.nesc.ac.uk:/cvsroot/pygmt co PyGMT
\end{verbatim}}
\texttt{PyGMT} is installed using
{\small
\begin{verbatim}
python setup.py install --home=$GLIMMER_PREFIX
\end{verbatim}}

The visualisation tools are part of the GLIMMER CVS repository. You can get them using
{\small
\begin{verbatim}
cvs -d:pserver:anonymous@forge.nesc.ac.uk:/cvsroot/glimmer login
cvs -z3 -d:pserver:anonymous@forge.nesc.ac.uk:/cvsroot/glimmer co PyCF
\end{verbatim}}
and install them
{\small
\begin{verbatim}
python setup.py install --home=$GLIMMER_PREFIX
\end{verbatim}}
